---
title: "Useful Functions for R Data Management"
author: "Jacob Yousif"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Useful Libraries
```{r}
library(tidyverse) #Does a lot of stuff that makes data outputs neater
library(janitor) #Helps clean variable names
library(naniar) #Helps manage NAs
library(lubridate) #Standardizes date format
library(shiny) #Useful for creating apps
```

```{r}
library(ggplot2) #Quickly make/edit charts and graphs. Included with tidyverse
library(RColorBrewer) #Adds color palettes to ggplot
library(paletteer) #Adds custom colors to ggplot
library(ggthemes) #Adds plot themes to ggplot
```

```{r}
#This is just a practice dataset
library(palmerpenguins)
```

```{r}
data <- palmerpenguins::penguins #Sets up data as a practice dataset
```
Note: many of the commands in this doc will not run properly. This is because the data has been messed up by all the previous commands and/or the commands have nonexistent variable names/parametes for demo purposes.

##Examining Data Tables
#Bring datasets into R
```{r}
read_csv("filepath") #Finds a CSV file specified by "filepath" and loads it into R as a data frame
data.frame(data) #Creates a data frame from the values in data (data can be a vector, matrix, or subset of a data frame)
```

#Getting an idea of a dataset's structure
```{r}
data #Prints whole dataset
data$species #Prints just the column specified by $ (here, it is "species")
dim(data) #Prints "<number of rows> <number of columns>"
colnames(data) #Prints column names
names(data) #Same as colnames()
glimpse(data) #Prints rows, columns, colnames and types, and the first few rows of each column
summary(data) #Prints colnames and shows statistical overview for each column's contents (min/max values, quartiles, median, mean, number of NAs)

any_na(data) #Returns TRUE if there are NAs in a dataset, and FALSE if there are not
  # Note: Is from naniar. Has equivalent `anyNA` in base; not sure of the difference
n_miss(data) #Returns the number of total NAs in a dataset
miss_var_summary(data) #Displays how many NAs per variable, as well as percentage of variable that is made up of NAs

table(data$species) #Returns a matrix that shows how many of each  discrete value are represented in a vector
tabyl(data$species) #Same as table(), but returns in a fancy data frame
#Note: can do table/yl(data$species) or (data, species). The latter returns a nicer variable name.
```

##Manipulating Data
#Cleaning up data
```{r}
clean_names(data) #Returns entire dataset with cleaned-up names. Changes all names to lowercase and replaces all special characters with "_". 
  #Remember to do data <- clean_names(data); it won't actually affect the dataset unless the data are replaced. 
  #Also note that the return is an entire data table, not just the colnames.
colnames(data) <- c("name1", "name2", "...") #Assigns the values of names to be the column names of (data)

na_if(data, "-999") #If a value matches the value in quotes, replace it with an NA
```
#Formatting dates
```{r}
dmy(date) #Converts date to standard format Y-M-D which R recofgnizes as a date. Use dmy() if date is currently in dmy format, use mdy() if date is in mdy format, use ydm if in ydm format, etc.
dmy_hms(datetime) #Same as above but includes hour/minute/second. Outputs in Y-M-D, H:M:S UTC
```

#Selecting data subsets
```{r}
select(data, species, sex, bill_length_mm) #Selects columns
  #Note: can also be used to sort columns; will output columns   in order listed
filter(data, sex == "male") #Selects rows based on whether the variable(s) match the selection criteria
table(data$species) #Returns a matrix that shows how many of each  discrete value are represented in a vector
tabyl(data$species) #Same as table(), but returns in a fancy data frame
group_by(data$species) #Assigns an invisible number to categories in a dataset which then allows other functions to perform actions by category rather than by individual entry. (e.g. male = 0, female = 1, then summarize(avg = mean(height)) will give all male height and all female height). Can group by multiple variables.
```

#Creating data structures
```{r}
c("value1, value2, value3") #Creates a vector from data
data.frame(data) #Creates a data frame from a dataset (vector, matrix, or existing dataframe)
```

#Changing entries
```{r}
mutate(newcol = val) #Creates new col with values specified by val
  case_when() #Assigns categorical value to data based on whether it matches the expression found in ()
    #Syntax: case_when(x < 5.0 & x > 0.0 ~ "value")
  
pivot_longer(data, names_to=x, values_to=x) #Pivots data to "long" format. Creates new col with values set to the previous variable names.
pivot_wider(data, names_from=x, values_from=x) #Pivots data to "wide" format. Uses names_from as names for new cols, and moves values from values_from into cols specified by names_from value. 
#
#   Wide:             Long:
#   name a b c        name letter value
#   A    1 2 3        A    a      1
#   B    1 2 3        A    a      1
#   C    1 2 3        A    a      1
#                     A    b      2   etc.

separate()

x_join(data1, data2, ...) #Joins multiple datasets in manner specified by x
  #inner: Keeps entries that appear in all datasets
  #left: Keeps entries that appear in first listed dataset
  #right: Keeps entries that appear in last listed dataset
  #full: Keeps all entries
  #anti: Keeps entries that do not appear in all datasets (opposite of inner)
```

##Analysis of Data
#Creating analysis figures
```{r}
summarize(data, sum1 = function1(value1), sum2 = function2(value2)) #Creates a data frame containing a summary of functions. Computes function1 on all variables in value1, then stores the result in sum1, and does that for each subsequent variable.
  n() #Gives number of entries in a set
  n_distinct() #Gives number of distinct variables in a set
  across(c(values), function()) #Computes function() for each variable in c(values)
```

#Statistics
```{r}
mean() #Arithmetic mean
stats::sd() #Standard deviation
range() #Min and max values
```

#Charts and graphs
```{r}
hist(data$variable) #Automatically creates a histogram (bar chart) showing values in a vector

ggplot(data, aes(aes=vars...)) + layer_xx() #Creates plots via layering
  aes() #Sets x- and y-axes
    x = #sets x-axis
    y = #Sets y-axis
    fill = #Automatically sets fill color by variable given
    color = #Automatically sets primary color by variable (or specify a "color")
    group = 
#Note: format is geom() is its own object being added to the plot object. This also means can add multiple geoms to same plot (e.g. point()+smooth() for scatterplot with trendline)
  geom_x(na.rm=) #Changes geometry of the plot to be x
    point #scatter plot
    jitter #scatter plot with extra noise to variables to space points more evenly
    boxplot #boxes w/ error
    smooth #line of best fit
      (method=x) #change model of trendline
        lm #fit to linear model (straight line)
    line #connect-the-dots line
    bar #bar graph. lists # entries for each x-axis variable. no y-axis
      (stat=identity) #allows y-axis to be specified
      (position="dodge") #unstacks bars separated by fill
    col #bar graph with specific x- and y-axes
    histogram #histogram (groups continuous variables based on value range). no y-axis
      (bins=n) #specifies # of groups to use
    density #like histogram, but infinite bins. creates smooth curve. no y-axis
  scale_x/y_scalefactor() #scales axis based on scalefactor without changing original data. select x or y, and replace scalefactor with whatever the scale facor should be (e.g. sqrt, log10)
  facet_x() #Prints a series of plots separated by a specified variable
    wrap(~a) #Prints plots side-by-side
    grid(a~b) #Makes a grid of plots sorted by x=a, y=b
  labs(title="title", x="x-axis", y="y-axis") #Changes data labels
  theme() #Changes text in plot
    axis.text.x/y = element_text() #Changes text on axes as specified by element_text()
  scale_color_brewer() #Adds point color from RColorBrewer
  scale_fill_brewer() #Adds fill color from RColorBrewer
  scale_fill_manual(values = ) #Adds custom fill color via paletteer (palette must be assigned to a variable)
  xlim/ylim(lim1, lim2) #Sets x/y bounds of plot to be lim1/lim2
  #Options for any of these include:
    color = #Changes line color/primary color
    fill =  #Changes fill color
    alpha = #Changes opacity
```

#Maps
```{r}
ggmap(get_map(...)) + layer_xx #Prints a map via layering
  get_map(bbox, maptype = "type", source = "source") #Creates the map
    #bbox:
    make_bbox(c(xmin, xmax), c(ymin, ymax), f=x) #Sets bounding box for what part of map to use.
              #f: margin around data range
    #maptype: sets the background image used as the map
    #source: specifies a source to get maptype from
  geom_point(data = data, aes(x_var, y_var)) #Same function as ggplot with additional params
  labs() #same as ggplot
```

##Apps
#Shiny
```{r}
#Basic shiny app structure:
#####
library(shiny)

ui <- fluidPage() #Note: elements appear in order listed

server <- function(input, output, session) {}

shinyApp(ui, server)
######
fluidPage(elements) #Sets up the UI of the page. `elements` include:
  selectInput("id", "label", choices = c(...), selected = "var1") #Creates an input selection box
  #id: variable id of selection
  #label: Text label above selection box
  #choices: Vector containing different selection options
  #selected: Option that is selected by default
  radioButtons("id", "label", choices = c(...))
  titlePanel("Title") #Adds title to page
  helpText("text") #Adds grey text block
###
  function()
```

##Uncategorized
```{r}
count()
tally()
is.na(data) #Logically evaluates each value. Returns T if value == NA, F if value has a value.
list.files(path = "filepath", pattern = "x", full.names = T/F) #Lists all files in a folder. Outputs a list
  #pattern: string to search for in file name
  #full.names: Bool; if true, prints full filepath (from working directory)
lapply(data, function) #Applies function to each element in data (data must be a list)
```

##Lists
```{r}
list[[n]] #Returns n-most element in list (e.g. n=3, returns 3rd element)
strsplit(list, split = "x") #Removes "x" from end of each element
```


##Useful Links
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [shiny cheatsheet](https://shiny.rstudio.com/images/shiny-cheatsheet.pdf)

