---
title: "Lab 2 Homework"
author: "Jacob Yousif"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

## Instructions

Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!

1.  What is a vector in R?

    A vector is a **one-dimensional** list of data values. Data in vectors is organized in a linear fashion. It is a sequential list, with each data point in the list possessing one "x-value".

2.  What is a data matrix in R?

    A data matrix is a **two-dimensional** list of data values, where each data point is assigned an "x-value" and a "y-value", i.e. a place in a row and column. Data matrices can store individual row and column names as well.

3.  Below are data collected by three scientists (Jill, Steve, Susan in order) measuring temperatures of eight hot springs. Run this code chunk to create the vectors.

```{r}
spring_1 <- c(36.25, 35.40, 35.30)
spring_2 <- c(35.15, 35.35, 33.35)
spring_3 <- c(30.70, 29.65, 29.20)
spring_4 <- c(39.70, 40.05, 38.65)
spring_5 <- c(31.85, 31.40, 29.30)
spring_6 <- c(30.20, 30.65, 29.75)
spring_7 <- c(32.90, 32.50, 32.80)
spring_8 <- c(36.80, 36.45, 33.15)
```

4.  Build a data matrix that has the springs as rows and the columns as scientists.

```{r}
spring_data <- c(spring_1, spring_2, spring_3, spring_4, spring_5, spring_6, spring_7, spring_8) #c all spring data into one vector

spring_matrix <- matrix(spring_data, nrow = 8, ncol = 3) #build matrix & store in spring_matrix
```


5.  The names of the springs are 1.Bluebell Spring, 2.Opal Spring, 3.Riverside Spring, 4.Too Hot Spring, 5.Mystery Spring, 6.Emerald Spring, 7.Black Spring, 8.Pearl Spring. Name the rows and columns in the data matrix. Start by making two new vectors with the names, then use `colnames()` and `rownames()` to name the columns and rows.

```{r}
spring_rows <- c("Bluebell Spring", "Opal Spring", "Riverside Spring", "Too Hot Spring", "Mystery Spring", "Emerald Spring", "Black Spring", "Pearl Spring")
spring_cols <- c("Jill", "Steve", "Susan")

rownames(spring_matrix) <- spring_rows
colnames(spring_matrix) <- spring_cols
spring_matrix
```


6.  Calculate the mean temperature of all eight springs.

```{r}
meanlist <- rowMeans(spring_matrix) #better way
meanlist
```

7.  Add this as a new column in the data matrix.

```{r}
spring_matrix_with_mean <- cbind(spring_matrix, meanlist)
colnames(spring_matrix_with_mean) <- c(colnames(spring_matrix), "Average") #adds "Average" title to column 4
spring_matrix_with_mean
```

8.  Show Susan's value for Opal Spring only.

```{r}
spring_matrix[2,3]
```

9.  Calculate the mean for Jill's column only.

```{r}
mean(spring_matrix[,1])
```

10. Use the data matrix to perform one calculation or operation of your interest.

```{r}
#Adds a row for average of the measurements taken by each scientist

sci_meanlist <- colMeans(spring_matrix_with_mean) #gets means of columns
spring_matrix_with_doublemean <- rbind(spring_matrix_with_mean, sci_meanlist) #adds as column
rownames(spring_matrix_with_doublemean) = c(rownames(spring_matrix_with_mean), "Average") #adds name to new column

spring_matrix_with_doublemean
```

## Push your final code to GitHub!

Please be sure that you check the `keep md` file in the knit preferences.
