---
title: "`summarize()`, `tabyl()`, and `group_by()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals

*At the end of this exercise, you will be able to:*\
1. Use a combination of `select()`, `filter()`, and `mutate()` to transform data frames.\
2. Use the `skimr` package to produce summaries of data.\
3. Produce clean summaries of data using `summarize()`.\
4. Use `group_by()` in combination with `summarize()` to produce grouped summaries of data.

## Review

At this point, you should be comfortable using the functions of `dplyr`. If you need extra help, please [email me](mailto:%20jmledford@ucdavis.edu).

## Load the tidyverse and janitor

```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Install `skimr`

```{r}
#install.packages("skimr")
library("skimr")
```

## Load the data

For this lab, we will use the built-in data on mammal sleep patterns. From: *V. M. Savage and G. B. West. A quantitative, theoretical framework for understanding mammalian sleep. Proceedings of the National Academy of Sciences, 104 (3):1051-1056, 2007*.

```{r}
?msleep
names(msleep)
glimpse(msleep)
```

We will also use the awesome [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/articles/intro.html) data in the second part of lab.

```{r}
library("palmerpenguins")
palmerpenguins::penguins_raw
```

## dplyr Practice

Let's do a bit more practice to make sure that we understand `select()`, `filter()`, and `mutate()`. Start by building a new data frame `msleep24` from the `msleep` data that: contains the `name` and `vore` variables along with a new column called `sleep_total_24` which is the amount of time a species sleeps expressed as a proportion of a 24-hour day. Remove any rows with NA's and restrict the `sleep_total_24` values to less than 0.3. Arrange the output in descending order.

```{r}
msleep24 <- msleep %>%
  mutate(sleep_total_24 = sleep_total/24) %>% 
  select(name, vore, sleep_total_24) %>% 
  filter(sleep_total_24 >= 0.3) %>% 
  filter(!is.na(vore)) %>% 
  arrange(desc(sleep_total_24))
```

Did `dplyr` do what we expected? How do we check our output? Remember, just because your code runs it doesn't mean that it did what you intended.

```{r}
summary(msleep24)
```

Try out the new function `skim()` as part of the `skimr` package.

```{r}
skim(msleep24)
```

Histograms are also a quick way to check the output.

```{r}
hist(msleep24$sleep_total_24)
```

Don't forget we can also use `tabyl()` across one or many variables.

```{r}
msleep24 %>% 
  tabyl(vore) %>% 
  adorn_pct_formatting(digits=1)
```

## Practice

1.  Which taxonomic orders have species that belong to more than one class of `vore`?

```{r}
msleep %>% 
  select(order, name, vore) %>% 
  arrange(order) %>% 
  tabyl(order, vore)
```

## `summarize()`

`summarize()` will produce summary statistics for a given variable in a data frame. For example, if you are asked to calculate the mean of `sleep_total` for large and small mammals you could do this using a combination of commands, but it isn't very efficient or clean. We can do better!

```{r}
head(msleep)
```

For example, if we define "large" as having a `bodywt` greater than 200 then we get the following:

```{r}
large <- 
  msleep %>% 
  select(name, genus, bodywt, sleep_total) %>% 
  filter(bodywt > 200) %>% 
  arrange(desc(bodywt))
large
```

```{r}
mean(large$sleep_total)
```

We can accomplish the same task using the `summarize()` function to make things cleaner.

```{r}
msleep %>% 
  filter(bodywt > 200) %>% #large mammals
  summarize(mean_sleep_lg = mean(sleep_total))
```

-   Note: `summarize()` is different than `summary()`

You can also combine functions to make useful summaries for multiple variables.

```{r}
msleep %>% 
    filter(bodywt > 200) %>% 
    summarize(mean_sleep_lg = mean(sleep_total), 
              min_sleep_lg = min(sleep_total),
              max_sleep_lg = max(sleep_total),
              total = n())
```

## Practice

1.  What is the mean, min, and max `bodywt` for the taxonomic order Primates? Provide the total number of observations.
```{r}
msleep$order
```

```{r}
msleep %>% 
  select(order, name, bodywt) %>% 
  filter(order == "Primates") %>% 
  summarize(min_bodywt = min(bodywt), 
            max_bodywt = max(bodywt), 
            mean_bodywt = mean(bodywt), 
            total = n())
```

`n_distinct()` is a very handy way of cleanly presenting the number of distinct observations. Here we show the number of distinct genera over 100 in body weight.

```{r}
msleep %>% 
  filter(bodywt > 100) %>% 
  summarise(n_genera=n_distinct(genus))
```

There are many other useful summary statistics, depending on your needs: sd(), min(), max(), median(), sum(), n() (returns the length of a column), first() (returns first value in a column), last() (returns last value in a column) and n_distinct() (number of distinct values in a column).

## Practice

1.  How many genera are represented in the msleep data frame?

```{r}
summarize(msleep, n_genera = n_distinct(genus))
```

2.  What are the min, max, and mean `sleep_total` for all of the mammals? Be sure to include the total n.

```{r}
summarize(msleep, min_sleep = min(sleep_total), max_sleep = max(sleep_total), mean_sleep = mean(sleep_total), total = n())
```

## `group_by()`

The `summarize()` function is most useful when used in conjunction with `group_by()`. Although producing a summary of body weight for all of the mammals in the data set is helpful, what if we were interested in body weight by feeding ecology?

```{r}
msleep %>%
  group_by(vore) %>% #we are grouping by feeding ecology, a categorical variable
  summarize(min_bodywt = min(bodywt),
            max_bodywt = max(bodywt),
            mean_bodywt = mean(bodywt),
            total=n())
```

Note: group_by(msleep, vore) does nothing on its own; it merely assigns invisible groups to variables within the dataset which can be used by other functions to perform actions by group rather than by individual datapoint.

## Practice

1.  Calculate mean brain weight by taxonomic order in the msleep data.

```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brainwt = mean(brainwt))
```

2.  What does `NA` mean? How are NA's being treated by the summarize function?

```{r}
#I think an NA datapoint makes the whole mean become NA
```

3.  Try running the code again, but this time add `na.rm=TRUE`. What is the problem with Cetacea? Compare this to Carnivora.

```{r}
msleep %>% 
  group_by(order) %>% 
  summarize(mean_brainwt = mean(brainwt, na.rm=T))
#Result for Cetacea has become NaN, Carnivora is now fine
```
```{r}
msleep %>% 
  select(order, brainwt) %>% 
  filter(order == "Cetacea")
#Cetacea is NaN because there are no datapoints for cetacean brain weight
msleep %>% 
  select(order, brainwt) %>% 
  filter(order == "Carnivora")
#Meanwhile, there are data for Carnivora brainwt, so the NAs can be ignored and still return a result

#Note: NaN is different than NA. Rather than returning a result which is a blank space (NA), the function said "no, I can't do that" and returned no result (NaN).
```

## That's it! Let's take a break and then move on to part 2!

--\>[Home](https://jmledford3115.github.io/datascibiol/)
